Fractional Types

### 2019-11-23 fracGC/p.tex
final pass

By lifting the scoping restriction, the development in the previous
sections is already more general than the state of the art in ancilla
management.  It however shares the same limitation of requiring a
runtime check to ensure ancillae values are properly restored to their
allocation~\cite{10.1007/978-3-319-20860-2_13,Green:2013:QSQ:2491956.2462177}.
In this section, we address this limitation using a combination of
pointed types, singleton types, monads, and comonads.

@@

The toy language has two types (natural numbers and booleans) and two
functions (and their compositions). Say we wanted to prove that
\AgdaInductiveConstructor{compose isZero square} always returns
\AgdaInductiveConstructor{false} when applied to a non-zero natural
number. We can certainly do this proof in Agda (i.e., in the
meta-language of our formalization) but we would like to do the proof
within the toy language itself. The most important
reason is that it can then be used within the language to
optimize programs (or, for the full language, to
remove a runtime check).

@@

We can track concrete values as the first two tests
show. More interesting is \AgdaFunction{test3} which shows a
property that holds for all natural numbers $n$. Evaluation
proceeds ``symbolically'' in the type system. From the definition
of \Afun{eval}, it is clear that
\Afun{eval}~\Acon{square}~(\Acon{suc}~\Avar{n}) produces
(\Acon{suc}~\Avar{n}) * (\Acon{suc}~\Avar{n}) which by definition of
multiplication is an expression with a leading \Acon{suc} constructor
which is enough to determine that evaluating
\Acon{isZero} on it yields \Acon{false}. This form of partial
evaluation will prove to be quite expressive, allowing us to keep track
of ancillae values throughout complex programs as demonstrated in
Sec.~\ref{sec:cat}.

@@

@@

By lifting programs and their evaluation to the type level, we can
naturally leverage the typechecking process to verify properties of
interest, including the safe de-allocation of ancillae. One ``could''
just write programs in that style in $\Pi/\bullet$: indeed the dual nature of proofs
and programs is more and more exploited in languages like the one used
to formalize this paper (Agda). The examples in Sec.~\ref{sec:cat} are
written in that style and they are relatively similar to programs
in the ``normal'' $\Pi/D$ language.

However, it is also often the case than constructive proofs are
further processed to extract a native efficient programs that eschew
the overhead of maintaining information needed just for proof
invariants. In our case, the question is whether we can extract from
$\Pi/\bullet$ program, a program in $\Pi/D$ that uses a simpler type
system, a simpler runtime representation, and yet is guaranteed to be
safe and hence can run without the runtime checks associated with
de-allocation sites. In this section, we show that this indeed the
case.

@@

The function \Afun{Ext𝕌} maps a $\Pi/\bullet$ type to a $\Pi/D$ type
and a value in the type. The function \Afun{Ext∙⟶} maps a
$\Pi/\bullet$ combinator to a $\Pi/D$ combinator. And finally the
function \Afun{Ext≡} asserts that extracted code cannot throw an
exception (it must return \Acon{just}\ldots).

Each of these functions has one or two enlightening cases which we
explain:

In $\Pi/D$ the fractional type expresses that it expects a particular
value but lacks any mechanisms to enforce this requirement. Thus we
have no choice when mapping a fractional type from $\Pi/\bullet$ to
$\Pi/D$ than to just use this \Acon{𝟙/}~\Avar{v} type:

@@

When mapping $\Pi/\bullet$ combinators to
$\Pi/D$ combinators, the main interesting cases are for \Acon{η} and
\Acon{ε}. In those we just use the values from the pointed type as
choices for the ancilla value and expectation for the gc process:

@@

In this case, the singleton type in $\Pi/\bullet$ guarantees that the
runtime check cannot fail.

@@

In addition to the examples presented throughout the paper, we present
examples that provides novel perspectives on gc processes, first as
space transformers, and second as tracing operators. Both ideas are
inspired from known constructions in compact closed
categories~\cite{dblp:conf/calco/abramsky05}.

@@

There has been ample work on ancillae before us.  For example
\cite{PhysRevA.94.042337} explicitly represents the lifetime of
quantum wires, enabling quantum wires to be recycled, leading
to more efficient circuits.  Interesting applications of
ancillae is in the recent paper~\cite{DBLP:journals/cal/ZhouB19}
on Quantum runtime assertions.  In section~\ref{sub:id}, we
already mentioned dynamic assertion of entanglement; they
also show how to provide (dynamic) assertions on classical
states, and testing of superposition states.  Our techniques
should combine with theirs to provide even more possibilities
for dynamic assertions. Ancilla management is a complex
business, and the verified circuit compiler
ReVerC~\cite{amy2017verified} goes to great pains to
ensure that ancilla are dealt with properly.  Both
QCL~\cite{omer2002procedural} and Quipper~\cite{Green:2013:QSQ:2491956.2462177}
also do automatic (but not certified) ancilla management during code
generation.  This remains an active area of research, as
outlined in the CACM paper~\cite{valiron2015programming}.

Another interesting avenue is to understand how to merge
pointed $\Pi$ with another simple language for quantum
computing: QWIRE~\cite{Paykin:2017:QCL:3009837.3009894}.
QWIRE has many desirable meta-properties that have been
formally proven to hold.  Our respective paradigms appear eminently compatible,
however their language has an unguarded \texttt{discard} directive that
may lead to erroneous circuits. For example, their circuit
reversing operation is partial, whereas ours is provably total.

@@

Another different thread of work stems from categorical models
of quantum computation, especially those based on dagger
compact closed categories as introduced
in~\cite{Abramsky:2004:CSQ:1018438.1021878}, and greatly
expanded upon in~\cite{abramsky2009categorical}, or more
recently in~\cite{DBLP:conf/lics/HuotS19}. Various
results can be obtained in that general setting, such as the
no-cloning theorem~\cite{abramsky2009no}. The $2016$ book
\cite{chiribella2016quantum} is a good survey of the
developments since. Selinger~\cite{selinger2007dagger,Selinger:2011:FDH:1942319.1942398}
has been particularly active in this area. Of particular
note for our applications are two 2005 papers by Abramsky
\cite{abramsky2005structural,dblp:conf/calco/abramsky05}
with explicit examples of circuits using trace that are very useful tests
for our approach.

Also related are various approaches to quantum lambda
calculi, such as~\cite{dblp:journals/mscs/selingerv06,vantonder:2004,dowek2017lineal}.

>>>>

By lifting the scoping restriction, the development in the previous
sections is already more general than the state of the art in ancilla
management.  It however shares the same limitation of requiring a
runtime check to ensure ancillae values are properly restored to their
allocation
value~\cite{10.1007/978-3-319-20860-2_13,Green:2013:QSQ:2491956.2462177}.
In this section, we address this limitation using a combination of
pointed types, singleton types, monads, and comonads.

@@

The toy language has two types (natural numbers and booleans) and two
functions (and their compositions). Say we wanted to prove that
\AgdaInductiveConstructor{compose isZero square} always returns
\AgdaInductiveConstructor{false} when applied to a non-zero natural
number. We can certainly do this proof in Agda (i.e., in the
meta-language of our formalization) but we would like to do the proof
within the toy language itself. The most important reason is that it
can then be used within the language to optimize programs (or, for the
case of $\Pi/D$ to remove a runtime check).

@@

We can track concrete values as the first two tests
show. More interesting is \AgdaFunction{test3} which shows a
property that holds for all natural numbers $n$. Evaluation
proceeds ``symbolically'' in the type system. From the definition
of \Afun{eval}, it is clear that
\Afun{eval}~\Acon{square}~(\Acon{suc}~\Avar{n}) produces
(\Acon{suc}~\Avar{n}) * (\Acon{suc}~\Avar{n}) which by definition of
multiplication is an expression with a leading \Acon{suc} constructor
which is enough to determine that evaluating
\Acon{isZero} on it yields \Acon{false}. This form of partial
evaluation will prove to be quite expressive, allowing us to keep track
of ancillae values throughout complex programs.

@@

As the trace shows, by garbage collecting the collector for value
$\mathbb{T}$, this value which was maintained in the type is now
``rematerialized.''

@@

By lifting programs and their evaluation to the type level, we can
naturally leverage the typechecking process to verify properties of
interest, including the safe de-allocation of ancillae. One ``could''
just forget about $\Pi/D$ and instead use $\Pi/\bullet$ as \emph{the}
programming language for ancilla management. Indeed the dual nature of
proofs and programs is more and more exploited in languages like the
one used to formalize this paper (Agda). The examples in
Sec.~\ref{sec:cat} are written in that style and they are relatively
similar to programs in the ``normal'' $\Pi/D$ language.

However, it is also often the case than constructive proofs are
further processed to extract native efficient programs that eschew
the overhead of maintaining information needed just for proof
invariants. In our case, the question is whether we can extract from
a $\Pi/\bullet$ program, a program in $\Pi/D$ that uses a simpler type
system, a simpler runtime representation, and yet is guaranteed to be
safe and hence can run without the runtime checks associated with
de-allocation sites. In this section, we show that this indeed the
case.

@@

\noindent The function \Afun{Ext𝕌} maps a $\Pi/\bullet$ type to a $\Pi/D$ type
and a value in the type. The function \Afun{Ext∙⟶} maps a
$\Pi/\bullet$ combinator to a $\Pi/D$ combinator. And finally the
function \Afun{Ext≡} asserts that extracted code cannot throw an
exception (it must return \Acon{just}\ldots).

Each of these functions has one or two enlightening cases which we
explain.

In $\Pi/D$ the fractional type expresses that it expects a particular
value but lacks any mechanisms to enforce this requirement. Thus we
have no choice when mapping a fractional type from $\Pi/\bullet$ to
$\Pi/D$ than to just use this \Acon{𝟙/}~\Avar{v} ~type:

@@

When mapping $\Pi/\bullet$ combinators to
$\Pi/D$ combinators, the main interesting cases are for \Acon{η} and
\Acon{ε}. In those, we just use the values from the pointed type as
choices for the ancilla value and expectation for the gc process:

@@

\noindent In this case, the singleton type in $\Pi/\bullet$ guarantees that the
runtime check cannot fail.

@@

In addition to the examples presented throughout the paper, we present
examples that provide novel perspectives on gc processes, first as
space transformers, and second as tracing operators. Both ideas are
inspired from known constructions in compact closed
categories~\cite{dblp:conf/calco/abramsky05}.

@@

We have introduced, in the context of reversible/quantum languages,
the concept of fractional types as descriptions of specialized gc
processes. Although the basic idea is rather simple and intuitive, the
technical details needed to reason about individual values are
somewhat intricate.

There has been ample work on ancillae before us.  For example
\cite{PhysRevA.94.042337} explicitly represents the lifetime of
quantum wires, enabling quantum wires to be recycled, leading
to more efficient circuits.  Interesting applications of
ancillae is in the recent paper~\cite{DBLP:journals/cal/ZhouB19}
on quantum runtime assertions.  In section~\ref{sub:id}, we
already mentioned dynamic assertion of entanglement; they
also show how to provide (dynamic) assertions on classical
states, and testing of superposition states.  Our techniques
should combine with theirs to provide even more possibilities
for dynamic assertions. Ancilla management is a complex
business, and the verified circuit compiler
ReVerC~\cite{amy2017verified} goes to great pains to
ensure that ancilla are dealt with properly.  Both
QCL~\cite{omer2002procedural} and Quipper~\cite{Green:2013:QSQ:2491956.2462177}
also do automatic (but not certified) ancilla management during code
generation.  This remains an active area of research, as
outlined in the CACM paper~\cite{valiron2015programming}.

Another interesting avenue is to understand how to merge $\Pi/\bullet$
with other languages for quantum
computing~\cite{dblp:journals/mscs/selingerv06,vantonder:2004,dowek2017lineal}. One
particular language, QWIRE~\cite{Paykin:2017:QCL:3009837.3009894}, has
a paradigm that appears eminently compatible to ours and has many
desirable meta-properties that have been formally proven to hold.
QWIRE however has an unguarded \texttt{discard} directive that may
lead to erroneous circuits. For example, their circuit reversing
operation is partial, whereas ours is provably total.

@@

Another different thread of work stems from categorical models
of quantum computation, especially those based on dagger
compact closed categories as introduced
in~\cite{Abramsky:2004:CSQ:1018438.1021878}, and greatly
expanded upon
in~\cite{abramsky2009categorical,selinger2007dagger,Selinger:2011:FDH:1942319.1942398}, or more
recently in~\cite{DBLP:conf/lics/HuotS19}. Various
results can be obtained in that general setting, such as the
no-cloning theorem~\cite{abramsky2009no}. The $2016$ book
\cite{chiribella2016quantum} is a good survey of the
developments since. Of particular
note for our applications are two 2005 papers by Abramsky
\cite{abramsky2005structural,dblp:conf/calco/abramsky05}
with explicit examples of circuits using trace that are very useful tests
for our approach.
